---
title: "Regression Models Course Project - Cars & MPG"
author: "gspinnadauhu"
date: "April 14, 2017"
output: pdf_document
---

##### Executive Summary

##### Data
This analysis is using the 'mtcars' dataset native to R.
Loading and examining the structure of the data:
```{r}
data("mtcars")
head(mtcars)
str(mtcars)
summary(mtcars)
```
This dataset has 32 observations and 11 variables.

There are no missing observations.
```{r}
sum(is.na(mtcars))
```
##### Questions
###### 1.“Is an automatic or manual transmission better for MPG”
First, plotting MPG by transmission (am: 0=auto, 1=manual)
```{r}
library(ggplot2)
ggplot(data=mtcars,aes(y=mpg,x=factor(am)))+
        geom_boxplot()+
        theme_classic()+
        labs(title="MPG by Transmission type",
             x="Transmission Type",
             y="Miles per Gallon")
```
It appears that manual transmission gets better mileage average overall, as the median falls above the upper quartile range of data for automatic transmission mpg.

Next, exploring the relationship between mpg and am, using a simple linear model.
```{r}
mpg_linear<-lm(mpg~am,mtcars)
summary(mpg_linear)
```

###### 2."Quantify the MPG difference between automatic and manual transmissions"

##### Conclusion