---
title: "Regression Models Course Project - Cars & MPG"
author: "gspinnadauhu"
date: "April 14, 2017"
output: pdf_document
---

##### Executive Summary

##### Data
This analysis is using the 'mtcars' dataset native to R.
Loading and examining the structure of the data:
```{r}
data("mtcars")
head(mtcars)
str(mtcars)
summary(mtcars)
```
This dataset has 32 observations and 11 variables.

There are no missing observations.
```{r}
sum(is.na(mtcars))
```
##### Questions
###### 1.“Is an automatic or manual transmission better for MPG”
First, plotting MPG by transmission (am: 0=auto, 1=manual)
```{r}
library(ggplot2)
ggplot(data=mtcars,aes(y=mpg,x=factor(am)))+
        geom_boxplot()+
        theme_classic()+
        labs(title="MPG by Transmission type",
             x="Transmission Type",
             y="Miles per Gallon")
```
It appears that manual transmission gets better mileage average overall, as the median falls above the upper quartile range of data for automatic transmission mpg.

Next, exploring the relationship between mpg (dependent variable) and am, using a simple linear model.
Let us first look at correllations between mpg and the other variables.
```{r}
cor(mtcars)[,1]
```
It appears here that mpg is highly correllated with #cylinders, engine displacement, horsepower, weight and to a lesser extent rear axle ratio.
These variables will be included, in addition to am, in the regression. Note that am (transmission) is not that highly correllated with mpg.

To avoid having too much multicollinearity, we now look at a correllation matrix between the chosen variables. Visualizing the correllations magnitudes in a correllation plot.
```{r}
library(corrplot)
corrplot(cor(mtcars[,-c(7,8,10,11)]),type="upper",tl.col="black",tl.srt=45)
```
Based on the above graphic, cyl, wt and disp are highly correllated between the other variables. This could lead to some multicollinearity when running the regression. We will make the determination if some of these should be excluded based on variance inflation factors.
It may also be useful to utilize interaction variables, i.e. hp/wt, which is a commonly used statistic for cars, usually called the "power to weight ratio".

The set of variables to be included in the linear model are the following:
y  mpg: miles per gallon (continuous)
x1 cyl: number of cylinders (discrete)
x2 disp: engine displacement (continuous)
x3 hp: horsepower (continuous)
x4 drat: real axle ratio (continuous)
x5 wt: weight in 1000lbs (continuous)
x6 am: Transmission type (0=automatic, 1=manual) (discrete)

```{r}
mpg_linear<-lm(mpg~factor(cyl)+disp+hp+drat+wt+factor(am),mtcars)
summary(mpg_linear)
```


Next, using sqrt of the variance inflations to see how much larger the standard error is due to inclusion of the variables.
```{r}
library(car)
sqrt(vif(mpg_linear))
```
Weight has a somewhat sd inflation factor, which was partially expected due to the high correllation with transmission type.

Next, looking at a residual plot.
```{r}
plot(predict(mpg_linear),resid(mpg_linear))
```
There appears to be no distinguishable pattern

###### 2."Quantify the MPG difference between automatic and manual transmissions"

##### Conclusion