---
title: "Regression Models Course Project - Cars & MPG"
author: "gspinnadauhu"
date: "April 14, 2017"
output: pdf_document
---

##### Executive Summary
This report assesses the effect of transmission type on gas mileage utilizing the Motor Trend dataset. A basic exploratory data analysis was performed, after which mpg distribution by transmission type was examined. In the second stage, based on correllation matrixes, independent variables were selected and an initial linear model was created. From there several statistical tests were ran and the model was adjusted based on their results. A final log-linear model was used to look at the effect of transmission type on mpg.
While, when looking at mpg solely by transmission type, manual transmissions appear to get better gas mileage, there was no measurable statistical relationship in our model. Confounding variables, like weight and hp, appear to have significant affect on mpg, while transmission type does not.

#### Questions
###### 1. “Is an automatic or manual transmission better for MPG”
Based on Figure 1 in the appendix, it would appear that cars with manual transmission get overall higher miles per gallon. However, once other variables are included, we cannot find a statistical significant relationship between transmission type and gas mileage.

###### 2. "Quantify the MPG difference between automatic and manual transmissions"
Since the transmission type coefficient in our models was not statistically significant, we cannot/must not interpret its value.
Comparing the means of cars with manual and with automatic transmission, we are able to see that they are statistically different from each other, but this relationship does not hold once confounding variables like, hp and weight, are introduced.

#### Appendix: Model Development
Loading and examining the structure of the data:

```{r results='hide'}
data("mtcars")
head(mtcars)
str(mtcars)
summary(mtcars)
```

First, plotting MPG by transmission (am: 0=auto, 1=manual)

```{r warning=FALSE, fig.height=2.5,fig.width=3.8}
library(ggplot2)
ggplot(data=mtcars,aes(y=mpg,x=factor(am)))+
        geom_boxplot()+
        theme_classic()+
        scale_x_discrete(labels=c("Automatic","Manual"))+
        labs(title="Figure 1: MPG by Transmission type",
             y="Miles per Gallon",
             x="")
```

It appears that manual transmission gets better mileage average overall, as the median falls above the upper quartile range of data for automatic transmission mpg.

We next test the differences in mean mpg for the transmission types with the following hypotheses.
H0: There is not difference in means of mpg
Ha: Mean MPG for Manual transmissions is less than for automatic transmission.

```{r warning=FALSE}
library(pander)
pander(with(mtcars,t.test(mpg[am==0],
                          mpg[am==1],
                          alternative="less",
                          paired=FALSE,
                          conf.level = 0.95)),
       caption = "T-Test Comparing Mean MPG for Manual and Automatic Transmission")
```

Based on the p-value of 0.0007, we reject the H0.
Next, exploring the relationship between mpg (dependent variable) and am, using a correllation matrix.

```{r results='hide'}
library(pander)
pander(cor(mtcars),
      caption="Correllation Coefficients with MPG")
```

It appears here that mpg is highly correllated with #cylinders, engine displacement, horsepower, weight and to a lesser extent rear axle ratio.
These variables will be included, in addition to am, in the regression. Note that am (transmission) is not that highly correllated with mpg.
The following model was set up to start the analysis:

```{r warning=FALSE}
library(pander)
mpg_linear<-lm(mpg~factor(cyl)+disp+hp+drat+wt+factor(am),mtcars)
pander(summary(mpg_linear),
       digits = 4,
       caption="Regression Output - Linear Model")
```

The adj R-sq suggests this model explains ~83% of variability, yet only hp's and wt's coefficients are significant at p<0.05.
Next, using variance inflation factors to see how addition of variables is inflating variance.

```{r warning=FALSE}
library(car)
library(pander)
pander(vif(mpg_linear),
       caption = "Variance Inflation Factors - Linear Model")
```

Displacement and cyl have very high factors as expected, as large engine displacement almost always comes from higher numbers of cylinders.
Since neither disp nor cyl had statistically significants coefficients, we will run a partial F-test to see if these can be excluded.
The partial F-test will be comparing an unrestricted model (mpg_linear) and the restricted model (mpg_linear less cyl and disp).
H0: Coefficients for both cyl and disp are zero.
Ha: At least one of their coefficients is not zero.

```{r warning=FALSE}
library(pander)
mpg_lm.rest<-lm(mpg~hp+drat+wt+factor(am),mtcars)
pander(anova(mpg_lm.rest,mpg_linear),
       caption = "Partial F-Test: Unrestricted vs Restricted Model")
```

Based on the p-value of 0.258, we fail to reject the null hypothesis. We therefore continue our analysis with the restricted model.
Looking at the variance inflation factors for the restricted model.

```{r warning=FALSE}
library(pander)
library(car)
pander(vif(mpg_lm.rest),
       caption="Variance Inflation Factors - Restricted Linear Model")
```

Based on the variance inflation factors, we can assume that there is only low-level of multicollinearity, as all factors are below 4.
Next, looking at a residual plot to see if there is heteroskedasticity or non-linearity in the model.

```{r warning=FALSE, message=FALSE, fig.height=2.5,fig.width=3.8}
library(ggplot2)
ggplot(data=mpg_lm.rest$model,aes(x=mpg_lm.rest$fitted.values,y=mpg_lm.rest$residuals))+
        geom_point()+
        labs(x="Fitted Values",y="Residuals",title="Figure 3: Residuals vs Fitted Values")+
        geom_smooth(aes(color="red"))+
        guides(color="none")+
        theme_classic()
```

There is potentially some non-linearity, but there is not a very clear pattern in this plot.
Testing for heteroskedasticity via the Breusch-Pagan Test, with H0: Constant error variance and Ha: Error variance changes with level of fitted values.

```{r warning=FALSE,results='hide'}
library(car)
library(pander)
ncvTest(mpg_lm.rest)
```

Based on the p-value of 0.114, we fail to reject the H0.
Now to see if the functional form for our model is appropriate, we use Ramsey's RESET test to check for model misspecification.
The test uses H0: coefficients for non-linear combinations of mpg (explanatory variable) are zero Ha: at least one of the coefficients is not zero.

```{r warning=FALSE,message=FALSE}
require(zoo)
library(lmtest)
library(pander)
pander(resettest(mpg_lm.rest),
       caption="RESET Test - Restricted Linear Model")
```

Based on the p-value of 0.022, we reject the H0 and conclude there that at least one of the coefficients for the non-linear combinations of mpg was not zero.
Setting up a log-linear model, by regressing log(mpg) onto the regressors and re-running the RESET Test on the new model.

```{r warning=FALSE}
library(lmtest)
library(pander)
mpg_loglin<-lm(log(mpg)~hp+drat+wt+factor(am),mtcars)
pander(resettest(mpg_loglin),
       caption = "RESET Test - Log-Linear Model")
```

Based on the RESET test's p-value of 0.0547, we now fail to reject the H0.
Below is a summary of our final model.

```{r warning=FALSE}
library(pander)
pander(summary(mpg_loglin),
       caption = "Regression Output - Log-Linear Model")
```

In terms of results, coefficients for hp and wt continue to be significant with the expected signs. Transmission type continues to be not significiant and its coefficient can therefor not be interpreted.